### 2016-10-02 paper/5_evaluation.tex
Minor fixes

In this chapter, we will investigate the scalability of Gilbert and its performance compared to famous hand-tuned ML algorithms.
We show that Gilbert is not only easily usable in terms of programming but also produces results with decent performance.

@@

\textbf{Matrix Multiplication}. 
As a first benchmark, we choose the matrix multiplication $A\times B$ with $A,B \in \mathbb{R}^{n\times n}$ with $n$ being the dimensionality. 
The matrix multiplication operation is demanding, both in CPU load as well as network I/O as it requires two network shuffles. 
The matrices $A$ and $B$ are sparse matrices with uniformly distributed non-zero cells. 
They are randomly generated prior to the matrix multiplication with a sparsity of $0.001$. 
As baseline, we run the same matrix multiplication on a single machine of the cluster using Gilbert's local execution engine. 
The Flink and Spark execution engines are both started with 48 gigabytes of main memory for their task managers and they distribute the work equally among all worker nodes. 
In the first experiment, we fixed the block sizes to $500 \times 500$ and set the number of cores to $64$. 
We then increased the dimensionality $n$ of $A$ and $B$ to observe the runtime behavior. 
The resulting execution times for the local, Flink and Spark execution engines are shown in \cref{fig:mmLoadRuntime}. 
In the second experiment, we investigate the scaling behavior of Gilbert with respect to the cluster size. 
In order to observe the inflicted communication costs, we keep the work load per core constant while increasing the number of cores. 
For this experiment we vary the number of cores from $1$ to $128$ and scale $n$ such that $n^3/\#cores$ is constant. 
We started with $n=5000$ for a single core and reached $n=25000$ on $128$ cores. 
The results of the experiment are shown in \cref{fig:mmNodesRuntime}.

On a single machine, we are able to execute the matrix multiplication for dimensionalities up to $n=10000$ in a reasonable time. 
We can see that the local execution performs better for dimensionalities $n \le 2500$. 
That is expected since the matrix still fits completely in the main memory of a single machine and the distributed execution adds some significant communication overhead and job start up latency. 
For matrices with $n>2500$, Spark and Flink start to calculate the matrix multiplication faster than the local executor. 
We also see that Gilbert can handle matrix sizes which scale far beyond what a single machine can handle. 
The results depicted in \cref{fig:mmNodesRuntime} indicate for both execution engines decent scale-out behavior.
For Spark we observe an almost horizontal line for number of cores $\le 32$.
For larger number of cores, the scaling degrades which is probably caused by spilling.
The same applies to Flink.
However, we can observe that it has to start spilling data earlier.
Thus, Flink's scale-out behaviour is not as good as Spark's with respect to matrix multiplication.

@@

\textbf{Gaussian Non-negative Matrix Factorization}. As second benchmark for evaluating the scalability properties, we choose the Gaussian non-negative matrix factorization (GNMF) algorithm~\cite{seung:anips2001a}.
GNMF finds for a given matrix $V \in \mathbb{R}^{d\times w}$ a factorization $W \in \mathbb{R}^{d\times t}$ and $H \in \mathbb{R}^{t\times w}$ such that $V\approx W H$ holds.
For our evaluation, we calculate one step of the GNMF algorithm.
We set $t=10$, $w=100000$ and vary the number of rows $d$ of $V$.
The matrix $V\in\mathbb{R}^{d\times 100000}$ is a sparse matrix whose sparsity is set to $0.001$.
The non-zero cells of $V$ are uniformly distributed.
As baseline, we run the GNMF on a single machine of the cluster using the local execution engine.
Like for the matrix multiplication benchmark, the task manager of Spark and Flink are started with 48 gigabytes of memory and both systems use the same scheduling strategy keeping the work load on all nodes equally distributed.
In the first experiment we fix the number of cores to $64$.
We start with $d=500$ and increase the number of rows of $V$ to $150000$.
The block size of Gilbert is set to $500 \times 500$.
In order to compare the results with the optimized NMF MapReduce implementation proposed in~\cite{liu:2010a}, we re-implemented the algorithm using the Spark and Flink runtime system.
This hand-tuned implementation, henceforth denoted as HT-GNMF (HT in the graphs), is also executed on $64$ cores.
The execution times of HT-GNMF and Gilbert's GNMF are shown in \cref{fig:nmfLoadRuntime}.
In the second experiment of the GNMF benchmark, we analyze how Gilbert scales-out when increasing the cluster size while keeping the work load for each core constant.
We vary the cluster size from $1$ core to $64$ cores and scale the number of documents $d$ accordingly.
Initially we start with $1000$ documents and, consequently, calculate the matrix factorization for $64000$ documents on $64$ cores.
The results of this experiment are shown in \cref{fig:nmfNodesRuntime}.

@@

			\addplot[
				color=teal,
				mark=triangle,
			]table[
				x=Rows,
				y=Time,
			]
			{data/nnmfStep/nnmfStepLoadFlink};
			\addplot[
				color=black,
				mark=diamond,
			]table[
				x=rowsV,
				y=time,
			]
			{data/nnmfStep/gnmfStepBenchLoadSpark};
			\addplot[
				color=magenta,
				mark=square,
			]table[
				x=rowsV,
				y=time,
			]
			{data/nnmfStep/gnmfStepBenchLoadFlink};
			\end{loglogaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:nmfLoadRuntime}
	\end{subfigure}
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Rows $d$ of $V$},
				ylabel={Speedup of HT-GNMF on Spark},
				width=\dualpgfwidth,
				ymin=0,
				ymax=2.5,
				legend entries={Spark,Flink},
				legend pos=north west,
			]
			\addplot[
				color=red,
				mark=o,
			] table[
				x=Rows,
				y=Speedup,
			]
			{data/nnmfStep/gnmfStepSpeedupSpark};

@@

The local executor can be applied to sizes of $d$ ranging from $500$ to $3000$ rows.
Surprisingly, the local execution is outperformed by the distributed engines.
This is likely related to Breeze's multi threading support.
The distributed systems can also be used for data sizes which cannot be handled by a single machine.
Both distributed execution engines scale well and achieve almost identical results as the hand tuned implementations.
The speedup of the HT-GNMF variants compared to GNMF on Spark's and Flink's executor is shown in \cref{fig:gnmfSpeedup}.
In the case of Flink, the HT-GNMF variant runs at most $1.78$ times faster than Gilbert's GNMF implementation for large data sets.
For Spark, we can only observe a maximum speedup of $1.35$.
This result underlines Gilbert's good performance.
Furthermore, the development with Gilbert was considerably easier.
One GNMF step can be programmed in five lines of Gilbert code, whereas we needed $28$ lines of Scala code for Spark's HT-GNMF and $70$ lines of Scala code for Flink's HT-GNMF.
Not only did we have to know how to program Spark and Flink, but it also took us quite some time to verify the correct functioning of both implementations.
The verification was made significantly more difficult and time-consuming due to a programming bug we introduced.
The debugging process showed us quite plainly how tedious the development process even with systems like Spark and Flink can be.
Thus, the productivity increase gained by using a high-level declarative programming language for linear algebra must not be neglected and compensates for the slight performance loss.
\cite{alvaro:2010a} made a similar observations while developing a declarative programming language for distributed systems programming.
The scale-out behavior of the Flink and Spark execution engines \cref{fig:nmfNodesRuntime} both show good results for $\#cores$ up to $64$.
The runtime on Spark with $64$ cores is only $1.26$ times slower than the runtime on a single core with the same work load per core.

@@

In this section, we want to investigate how well the PageRank algorithm~\cite{page:1999a} implemented in Gilbert performs compared to a specialized implementation (denoted as SP in the figures).
We expect that the Gilbert runtime adds some overhead.
Furthermore, the high-level linear algebra abstraction of Gilbert might make it difficult to exploit certain properties to speed up the processing.
We first execute PageRank directly in Gilbert, given the Gilbert code \cref{lst:gilbertPageRank}, and then run them directly on Flink and Spark.
In contrast to Gilbert, the direct implementation requires a deep understanding of the underlying runtime system.
Furthermore, the distributed implementation is far from trivial compared to the linear algebra representation of the original problem.

@@

The Spark and Flink implementation of PageRank works as follows.
The PageRank vector is represented by a set of tuples $(w_i, r_i)$ with $w_i$ denoting the web site $i$ and $r_i$ being its rank.
The adjacency matrix is stored row-wise as a set of tuples $(w_i, A_i)$ with $w_i$ denoting the web site $i$ and $A_i$ being its adjacency list.
For the adjacency list $A_i$, it holds that $w_j \in A_i$ if and only if there exists a link from web site $i$ to $j$.
The next PageRank vector is calculated as follows.
The PageRank $r_i$ of web site $i$ is joined with its adjacency list $A_i$.
For each outgoing link $w_j \in A_i$ a new tuple $(w_j, 0.85r_i/\left|A_i\right|)$ with the rank of $i$ being divided by the number of outgoing links is created.
In order to incorporate the random jump behavior to any available web site, a tuple $(w_i, 0.15/|w|)$, with $|w|$ being the number of all web sites, is generated for each web site $i$.
In order to compute the next PageRank vector, all newly generated tuples are grouped according to their ID.
The resulting groups are reduced by adding up all partial rank values giving the new PageRank vector.
These steps constitute a single PageRank iteration whose data flow plan is depicted in \cref{fig:pageRankDataFlow}.

@@

			\addplot[red,
				mark=o,
			] table[
				x=NumVertices,
				y=Speedup,
			]
			{data/pagerank/pagerankSpeedupFlink};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:pageRankSpeedup}
	\end{subfigure}
	\caption{Comparison of Gilbert's PageRank implementation with specialized algorithms on Spark and Flink running on $64$-core cluster. \subref{fig:pageRankResults} Execution time of $10$ steps of the PageRank algorithm depending on the adjacency matrix's size. \subref{fig:pageRankSpeedup} Speedup of specialized algorithms with respect to Gilbert's implementations.}
	\label{fig:pageRankEvaluation}
\end{figure}

The graphs show that the specialized PageRank algorithm runs faster than Gilbert's versions of PageRank \cref{fig:pageRankResults}.
Furhtermore, the specialized algorithms show a better scalability.
The speedup of the specialized algorithms with respect to Gilbert's implementations for this experiment is shown in \cref{fig:pageRankSpeedup}.
For $n\le 50000$, the hand-tuned algorithms running on Spark and Flink show a similarl speedup.
The Flink and Spark version achieve a speedup of approximately $14$ and $10$ for $n=50000$, respectively.
However, for $n = 100000$ we can observe a speedup of $76$ for the Flink specialized implementation.

>>>>

In this chapter, we will investigate the scalability of Gilbert and its performance compared to well known hand-tuned ML algorithms.
We show that Gilbert is not only easily usable in terms of programming but also produces results with decent performance.

@@

\textbf{Matrix Multiplication}. 
As a first benchmark, we choose the matrix multiplication $A\times B$ with $A,B \in \mathbb{R}^{n\times n}$ and $n$ being the dimensionality. 
The matrix multiplication operation is demanding, both in CPU load as well as network I/O as it requires two network shuffles. 
The matrices $A$ and $B$ are sparse matrices with uniformly distributed non-zero cells. 
They are randomly generated prior to the matrix multiplication with a sparsity of $0.001$. 
As baseline, we run the same matrix multiplication on a single machine of the cluster using Gilbert's local execution engine. 
The Flink and Spark execution engines are both started with 48 gigabytes of main memory for their task managers and they distribute the work equally among all worker nodes. 
In the first experiment, we fixed the block sizes to $500 \times 500$ and set the number of cores to $64$. 
We then increased the dimensionality $n$ of $A$ and $B$ to observe the runtime behavior. 
The resulting execution times for the local, Flink and Spark execution engines are shown in \cref{fig:mmLoadRuntime}. 
In the second experiment, we investigate the scaling behavior of Gilbert with respect to the cluster size. 
In order to observe the inflicted communication costs, we keep the work load per core constant while increasing the number of cores. 
For this experiment we vary the number of cores from $1$ to $128$ and scale $n$ such that $n^3/\#cores$ is constant. 
We started with $n=5000$ for a single core and reached $n=25000$ on $128$ cores. 
The results of the experiment are shown in \cref{fig:mmNodesRuntime}.

On a single machine, we are able to execute the matrix multiplication for dimensionalities up to $n=10000$ in a reasonable time. 
We can see that the local execution performs better for dimensionalities $n \le 2500$. 
That is expected since the matrix still fits completely in the main memory of a single machine and the distributed execution adds some significant communication overhead and job start up latency. 
For matrices with $n>2500$, Spark and Flink start to calculate the matrix multiplication faster than the local executor. 
We also see that Gilbert can handle matrix sizes which scale far beyond what a single machine can handle.

The results depicted in \cref{fig:mmNodesRuntime} indicate for both execution engines decent scale-out behavior.
For Spark we observe an almost horizontal line for number of cores $\le 32$.
For larger number of cores, the scaling degrades which is probably caused by spilling.
The same applies to Flink.
However, we can observe that the system has to start spilling data earlier.
Thus, Flink's scale-out behaviour is not as good as Spark's with respect to matrix multiplication.

@@

\textbf{Gaussian Non-negative Matrix Factorization}. As second benchmark for evaluating the scalability properties, we choose the Gaussian non-negative matrix factorization (GNMF) algorithm~\cite{seung:anips2001a}.
GNMF finds for a given matrix $V \in \mathbb{R}^{d\times w}$ a factorization $W \in \mathbb{R}^{d\times t}$ and $H \in \mathbb{R}^{t\times w}$ such that $V\approx W H$ holds.
For our evaluation, we calculate one step of the GNMF algorithm.
We set $t=10$, $w=100000$ and vary the number of rows $d$ of $V$.
The matrix $V\in\mathbb{R}^{d\times 100000}$ is a sparse matrix whose sparsity is set to $0.001$.
The non-zero cells of $V$ are uniformly distributed.
As baseline, we run the GNMF on a single machine of the cluster using the local execution engine.
Like for the matrix multiplication benchmark, the task manager of Spark and Flink are started with 48 gigabytes of memory.
In the first experiment we fix the number of cores to $64$.
We start with $d=500$ and increase the number of rows of $V$ to $150000$.
The block size of Gilbert is set to $500 \times 500$.
In order to compare the results with the optimized GNMF MapReduce implementation proposed in~\cite{liu:2010a}, we re-implemented the algorithm using the Spark and Flink runtime system.
This hand-tuned implementation, henceforth denoted as HT-GNMF (HT in the graphs), is also executed on $64$ cores.
The execution times of HT-GNMF and Gilbert's GNMF are shown in \cref{fig:nmfLoadRuntime}.
In the second experiment of the GNMF benchmark, we analyze how Gilbert scales-out when increasing the cluster size while keeping the work load for each core constant.
We vary the cluster size from $1$ core to $64$ cores and scale the number of documents $d$ accordingly.
Initially we start with $1000$ documents and, consequently, calculate the matrix factorization for $64000$ documents on $64$ cores.
The results of this experiment are shown in \cref{fig:nmfNodesRuntime}.

@@

			\addplot[
				color=teal,
				mark=triangle,
			]table[
				x=Rows,
				y=Time,
			]
			{data/nnmfStep/nnmfStepLoadFlink};
			\addplot[
				color=black,
				mark=diamond,
			]table[
				x=rowsV,
				y=time,
			]
			{data/nnmfStep/gnmfStepBenchLoadSpark};
			\addplot[
				color=magenta,
				mark=square,
			]table[
				x=rowsV,
				y=time,
			]
			{data/nnmfStep/gnmfStepBenchLoadFlink};
			\end{loglogaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:nmfLoadRuntime}
	\end{subfigure}
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Rows $d$ of $V$},
				ylabel={Speedup of HT-GNMF over Gilbert},
				width=\dualpgfwidth,
				ymin=0,
				ymax=2.5,
				legend entries={Spark,Flink},
				legend pos=north west,
			]
			\addplot[
				color=red,
				mark=o,
			] table[
				x=Rows,
				y=Speedup,
			]
			{data/nnmfStep/gnmfStepSpeedupSpark};

@@

The local executor is applied to sizes of $d$ ranging from $500$ to $3000$ rows.
Surprisingly, the local execution is outperformed by the distributed engines.
This fact indicates that the network communication costs are not decisive.
The distributed systems can also be used for data sizes which can no longer be handled by a single machine.
Both distributed execution engines scale well and achieve almost identical results as the hand tuned implementations.
The speedup of the HT-GNMF variants compared to GNMF on Spark's and Flink's executor is shown in \cref{fig:gnmfSpeedup}.
In the case of Flink, the HT-GNMF variant runs at most $1.78$ times faster than Gilbert's GNMF implementation for large data sets.
For Spark, we can only observe a maximum speedup of $1.35$.
This result underlines Gilbert's good performance.
Furthermore, the development with Gilbert was considerably easier.
One GNMF step can be programmed in five lines of Gilbert code, whereas we needed $28$ lines of Scala code for Spark's HT-GNMF and $70$ lines of Scala code for Flink's HT-GNMF.
Not only did we have to know how to program Spark and Flink, but it also took us quite some time to verify the correct functioning of both implementations.
The verification was made significantly more difficult and time-consuming due to a programming bug we introduced.
The debugging process showed us quite plainly how tedious the development process even with systems like Spark and Flink can be.
Thus, the productivity increase gained by using a high-level declarative programming language for linear algebra must not be neglected and compensates for the slight performance loss.
\cite{alvaro:2010a} made a similar observations while developing a declarative programming language for distributed systems programming.
The scale-out behavior of the Flink and Spark execution engines \cref{fig:nmfNodesRuntime} both show good results for $\#cores$ up to $64$.
The runtime on Spark with $64$ cores is only $1.26$ times slower than the runtime on a single core with the same work load per core.
For Flink we observe a slowdown of $2.38$ when running on $64$ cores.

@@

In this section, we want to investigate how well the PageRank algorithm~\cite{page:1999a} implemented in Gilbert performs compared to a specialized implementation (denoted as SP in the figures).
We expect that the Gilbert runtime adds some overhead.
Furthermore, the high-level linear algebra abstraction of Gilbert might make it difficult to exploit certain properties to speed up the processing.
We first execute PageRank directly in Gilbert, given the Gilbert code in \cref{lst:gilbertPageRank}, and then run them directly on Flink and Spark.
In contrast to Gilbert, the direct implementation requires a deep understanding of the underlying runtime system.
Furthermore, the distributed implementation is far from trivial compared to the linear algebra representation of the original problem.

@@

The specialized Spark and Flink implementation of PageRank works as follows.
The PageRank vector is represented by a set of tuples $(w_i, r_i)$ with $w_i$ denoting the web site $i$ and $r_i$ being its rank.
The adjacency matrix is stored row-wise as a set of tuples $(w_i, A_i)$ with $w_i$ denoting the web site $i$ and $A_i$ being its adjacency list.
For the adjacency list $A_i$, it holds that $w_j \in A_i$ if and only if there exists a link from web site $i$ to $j$.
The next PageRank vector is calculated as follows.
The PageRank $r_i$ of web site $i$ is joined with its adjacency list $A_i$.
For each outgoing link $w_j \in A_i$ a new tuple $(w_j, 0.85r_i/\left|A_i\right|)$ with the rank of $i$ being divided by the number of outgoing links is created.
In order to incorporate the random jump behavior to any available web site, a tuple $(w_i, 0.15/|w|)$, with $|w|$ being the number of all web sites, is generated for each web site $i$.
In order to compute the next PageRank vector, all newly generated tuples are grouped according to their ID and summed up.
These steps constitute a single PageRank iteration whose data flow plan is depicted in \cref{fig:pageRankDataFlow}.

@@

			\addplot[red,
				mark=o,
			] table[
				x=NumVertices,
				y=Speedup,
			]
			{data/pagerank/pagerankSpeedupFlink};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:pageRankSpeedup}
	\end{subfigure}
	\caption{Comparison of Gilbert's PageRank implementation with specialized algorithms on Spark and Flink. \subref{fig:pageRankResults} Execution time of $10$ steps of the PageRank algorithm depending on the adjacency matrix's size. \subref{fig:pageRankSpeedup} Speedup of specialized algorithms with respect to Gilbert's implementations.}
	\label{fig:pageRankEvaluation}
\end{figure}

The graphs in \cref{fig:pageRankResults} show that the specialized PageRank algorithm runs faster than Gilbert's versions of PageRank.
Furthermore, the specialized algorithms show a better scalability.
The speedup of the specialized algorithms with respect to Gilbert's implementations for this experiment is shown in \cref{fig:pageRankSpeedup}.
For $n\le 50000$, the hand-tuned algorithms running on Spark and Flink show a similar speedup.
The Flink and Spark version achieve a speedup of approximately $13$ and $12$ for $n=50000$, respectively.
However, for $n = 100000$ we can observe a speedup of $76$ for the Flink specialized implementation whereas Spark reaches a speedup of $14$.
Gilbert's performance degradation with Flink is likley caused by earlier data spilling.

### 2016-10-02 paper/6_related_work.tex
tilmanns changes + shortening
Minor fixes

%!TEX root=main.tex
\section{Related Work}
\label{sec:relatedWork}

{\em SystemML}~\cite{ghoting:2011a,Boehm2014,Boehm2014Hybrid,Schelter2015,Elgohary2016} aims to make machine learning algorithms run on massive datasets without burdening the user with low-level implementation details and tedious hand-tuning. Therefore, it provides an R-like declarative high-level language, called Declarative Machine learning Language (\emph{DML}), and compiles and optimizes the resulting programs to distributed dataflow systems.  The linear algebra operations are translated into a directed acyclic graph of high-level operators. This abstract representation allows to apply several optimizations such as algebraic rewrites, choice of internal matrix representation and cost-based selection of the best execution plan. Afterwards these plans are translated to low-level operators and executed either in the driver memory or distributed. SystemML is one of the main inspirations for Gilbert. While it originally only supported MapReduce as runtime, it has recently also moved to supporting more advanced dataflow systems. Gilbert differs from SystemML by its fixpointer operator and by leveraging the general optimizers of the underlying system (e.g., the Flink optimizer for matrix multiplication optimization, see~Section~\ref{sec:LinearAlgebraOperations}).

{\em Samsara}~\cite{Lyubimov2016, Schelter2016} is a domain-specific language for declarative machine learning in cluster environments. Samsara allows its users to specify programs using a set of common matrix abstractions and linear algebraic operations, similar to R or MATLAB. Samsara automatically compiles, optimizes and executes these programs on distributed dataflow systems. The aim of Samsara is to allow mathematicians and data scientists to leverage the scalability of distributed dataflow systems via common declarative abstractions, without the need for deep knowledge about the programming and execution models of such systems. Samsara is part of the Apache Mahout library~\cite{mahout:2011a} and supports a variety of backends, such as Apache Spark or Apache Flink. {\em Ricardo}~\cite{das:2010a} uses existing technologies to implement a scalable system for statistical analysis: it executes the data shipment via Hadoop and runs the actual analytic computation by R. It integrates these systems via Jaql~\cite{beyer:2011a}, a declarative high-level language for data-processing on Hadoop. Thereby, the user can initiate the distribution, transformation and aggregation of data within Hadoop. Furthermore, the system supports to run R code on the worker nodes as data transformations. However, in this setting the user still has to explicitly specify the data-flow and the data distribution, which requires a substantial understanding of the working principles of MapReduce. {\em RHIPE}~\cite{guha:s2012a} also tries to extend R to scale to large data sets using Hadoop. RHIPE follows an approach called divide and recombine, The idea is to split the examined data up into several chunks so that they fit in the memory of a single computer, next a collection of analytic methods is applied to each chunk without communicating with any other computer. After each chunk is evaluated, the results will be recombined in an aggregation step to create the overall analytic result. However, this strict execution pipeline constrains the analysis process considerably. A system, which integrates more seamlessly into the R ecosystem, is {\em pR} (parallel R)~\cite{samatova:2009a}. Again, the goal is to let the statistician compute large-scale data without requiring him to learn a new system. pR achieves its goal by providing a set of specialized libraries which offer parallelized versions of different algorithms (using MPI). A drawback of this approach is that nowadays, many data analysts share a cluster of commodity hardware and run several jobs concurrently, some of which are interactive and others take a long time to finish. This multi-job scenario interferes with the design of MPI, which furthermoreinherently lacks often requested properties such as fault-tolerance, elasticity and reliability. Another related system for R is {\em rmpi}~\cite{rmpi}. As this is only a wrapper for the respective MPI calls, it also suffers from the aforementioned problems. Furthermore, there is the SparkR~\cite{Venkataraman2016} project which aims to integrate Spark's API and SQL processing capabilities into R.

For the other popular numerical computing environment out there, namely MATLAB, also a set of parallelization tools exists. The most popular are the MATLAB Parallel Computing Toolbox~\cite{parallelComputingToolbox} and MATLAB Distributed Computing Server~\cite{distributedComputingServer}, which are developed by MathsWorks. The former permits to parallelize MATLAB code on a multi-core computer and the latter scales the parallelized code up to large cluster of computing machines. In combination, they offer a flexible way to specify parallelism in MATLAB code. Besides MATLAB Parallel Computing Toolbox there are also other projects which try to parallelize MATLAB code. The most noteworthy candidates are pMatlab~\cite{bliss:ijhpca2007a} and MatlabMPI~\cite{kepner:jpdc2004a} which shall be named here for the sake of completeness. Unfortunately, none of these approaches is known to integrate will with the current JVM-based Hadoop ecosystem, which is becoming the main source of data in production deployments.

Another promising research direction is the deep embedding of the APIs of dataflow systems in their host language~\cite{Alexandrov2015}, where the potential to extend this embedding and the resulting optimizations to linear algebraic operations is currently explored~\cite{Kunft2016}.

>>>>

%!TEX root=paper.tex
\section{Related Work}
\label{sec:relatedWork}

{\em SystemML}~\cite{ghoting:2011a,Boehm2014,Boehm2014Hybrid,Schelter2015,Elgohary2016} aims to make machine learning algorithms run on massive datasets without burdening the user with low-level implementation details and tedious hand-tuning. Therefore, it provides an R-like declarative high-level language, called Declarative Machine learning Language (\emph{DML}), and compiles and optimizes the resulting programs to distributed dataflow systems.  The linear algebra operations are translated into a directed acyclic graph of high-level operators. This abstract representation allows to apply several optimizations such as algebraic rewrites, choice of internal matrix representation and cost-based selection of the best execution plan. Afterwards these plans are translated to low-level operators and executed either in the driver memory or distributedly. SystemML is one of the main inspirations for Gilbert. While it originally only supported MapReduce as runtime, it has recently also moved to supporting more advanced dataflow systems. Gilbert differs from SystemML by its fixpoint operator and by leveraging the general optimizers of the underlying system (e.g., the Flink optimizer for matrix multiplication optimization, see~Section~\ref{sec:LinearAlgebraOperations}). {\em Samsara}~\cite{Lyubimov2016, Schelter2016} is a domain-specific language for declarative machine learning in cluster environments. Samsara allows its users to specify programs using a set of common matrix abstractions and linear algebraic operations, similar to R or MATLAB. Samsara automatically compiles, optimizes and executes these programs on distributed dataflow systems. The aim of Samsara is to allow mathematicians and data scientists to leverage the scalability of distributed dataflow systems via common declarative abstractions, without the need for deep knowledge about the programming and execution models of such systems. Samsara is part of the Apache Mahout library~\cite{mahout:2011a} and supports a variety of backends, such as Apache Spark or Apache Flink. {\em Ricardo}~\cite{das:2010a} uses existing technologies to implement a scalable system for statistical analysis: it executes the data shipment via Hadoop and runs the actual analytic computation by R. It integrates these systems via Jaql~\cite{beyer:2011a}, a declarative high-level language for data-processing on Hadoop. Thereby, the user can initiate the distribution, transformation and aggregation of data within Hadoop. Furthermore, the system supports to run R code on the worker nodes as data transformations. However, in this setting the user still has to explicitly specify the data-flow and the data distribution, which requires a substantial understanding of the working principles of MapReduce. {\em RHIPE}~\cite{guha:s2012a} also tries to extend R to scale to large data sets using Hadoop. RHIPE follows an approach called divide and recombine, The idea is to split the examined data up into several chunks so that they fit in the memory of a single computer. Next a collection of analytic methods is applied to each chunk without communicating with any other computer. After each chunk is evaluated, the results will be recombined in an aggregation step to create the overall analytic result. However, this strict execution pipeline constrains the analysis process considerably. A system, which integrates more seamlessly into the R ecosystem, is {\em pR} (parallel R)~\cite{samatova:2009a}. Again, the goal is to let the statistician compute large-scale data without requiring him to learn a new system. pR achieves its goal by providing a set of specialized libraries which offer parallelized versions of different algorithms (using MPI). A drawback of this approach is that nowadays, many data analysts share a cluster of commodity hardware and run several jobs concurrently, some of which are interactive and others take a long time to finish. This multi-job scenario interferes with the design of MPI, which furthermore inherently lacks often requested properties such as fault-tolerance, elasticity and reliability. Another related system for R is {\em rmpi}~\cite{rmpi}. As this is only a wrapper for the respective MPI calls, it also suffers from the aforementioned problems. Furthermore, there is the SparkR~\cite{Venkataraman2016} project which aims to integrate Spark's API and SQL processing capabilities into R. For the other popular numerical computing environment out there, namely MATLAB, also a set of parallelization tools exists. The most popular are the MATLAB Parallel Computing Toolbox~\cite{parallelComputingToolbox} and MATLAB Distributed Computing Server~\cite{distributedComputingServer}, which are developed by MathsWorks. The former permits to parallelize MATLAB code on a multi-core computer and the latter scales the parallelized code up to large cluster of computing machines. In combination, they offer a flexible way to specify parallelism in MATLAB code. Besides MATLAB Parallel Computing Toolbox there are also other projects which try to parallelize MATLAB code. The most noteworthy candidates are pMatlab~\cite{bliss:ijhpca2007a} and MatlabMPI~\cite{kepner:jpdc2004a} which shall be named here for the sake of completeness. Unfortunately, none of these approaches is known to integrate will with the current JVM-based Hadoop ecosystem, which is becoming the main source of data in production deployments. Another promising research direction is the deep embedding of the APIs of dataflow systems in their host language~\cite{Alexandrov2015}, where the potential to extend this embedding and the resulting optimizations to linear algebraic operations is currently explored~\cite{Kunft2016}.

### 2016-10-02 paper/7_conclusion.tex
Sebastian's shortening tasks
Minor fixes

Gilbert tackles the problem of scalable analytics by fusing the assets of high-level linear algebra abstractions with the power of massively parallel dataflow systems.
Gilbert is a fully functional linear algebra environment, which is programmed by the widespread MATLAB language.
Gilbert programs are executed on massively parallel dataflow systems, which allow to process data exceeding the capacity of a single computer's memory.

The system itself comprises the technology stack to parse, type and compile MATLAB code into a format which can be executed in parallel.
In order to apply high-level linear algebra optimizations, we conceived an intermediate representation for linear algebra programs.
Gilbert's optimizer exploits this representation to remove redundant transpose operations and to determine an optimal matrix multiplication order.
The optimized program is translated into an highly optimized execution plan suitable for the execution on a supported engine.
Gilbert allows the distributed execution on Spark and Flink.

Our systematical evaluation has shown that Gilbert supports all fundamental linear algebra operations, making it fully operational.
Additionally, its loop support allows to implement a wide multitude of machine learning algorithms within Gilbert.
Exemplary, we have implemented the PageRank and GNMF algorithm.
The code changes required to make these algorithms run in Gilbert are minimal and only concern Gilbert's loop abstraction.
Our benchmarks have proved that Gilbert can handle data sizes which no longer can be efficiently processed on a single computer.
Moreover, Gilbert showed a promising scale out behavior, making it a suitable candidate for large-scale data processing.

The key contributions of this work include the development of a scalable data analysis tool which can be programmed using the well-known MATLAB language.
We have introduced the fixed-point operator which allows to express loops in a recursive fashion.
The key characteristic of this abstraction is that it allows an efficient concurrent execution unlike the \code{for} and \code{while} loops.
Furthermore, we researched how to implement linear algebra operations efficiently in modern parallel data flow systems, such as Spark and Flink.
Last but not least, we have shown that Flink's optimizer is able to automatically choose the best execution strategy for matrix-matrix multiplications.

Even though, Gilbert can process vast amounts of data, it turned out that it cannot beat the performance of hand-tuned algorithms using Spark or Flink.
This is caused by the overhead entailed by the linear algebra abstraction.
The overhead is also the reason for the higher memory foot-print, causing Spark and Flink to spill faster to disk than the hand-tuned algorithms.
Gilbert trades some performance off for better usability, which manifests itself in shorter and more expressive programming code.
The fact that Gilbert only supports one hard-wired partitioning scheme, namely square blocks, omits possible optimization potential.

Interesting aspects for further research and improvements of Gilbert include the extension of the Gilbert optimizer by new optimization strategies.
The investigation of different matrix partitioning schemes and its integration into the optimizer which selects the best overall partitioning seems to be very promising.
Furthermore, a tighter coupling of Gilbert's optimizer with Flink's optimizer could result in beneficial synergies.
Forwarding the inferred matrix sizes to the underlying execution system might help to improve the execution plans.

>>>>

Gilbert tackles the problem of scalable analytics by fusing the assets of high-level linear algebra abstractions with the power of massively parallel dataflow systems. Gilbert is a fully functional linear algebra environment, which is programmed by the widespread MATLAB language. Gilbert programs are executed on massively parallel dataflow systems, which allow to process data exceeding the capacity of a single computer's memory. The system itself comprises the technology stack to parse, type and compile MATLAB code into a format which can be executed in parallel. In order to apply high-level linear algebra optimizations, we conceived an intermediate representation for linear algebra programs. Gilbert's optimizer exploits this representation to remove redundant transpose operations and to determine an optimal matrix multiplication order. The optimized program is translated into an highly optimized execution plan suitable for the execution on a supported engine. Gilbert allows the distributed execution on Spark and Flink. Our systematical evaluation has shown that Gilbert supports all fundamental linear algebra operations, making it fully operational.
Additionally, its loop support allows to implement a wide multitude of machine learning algorithms.
Exemplary, we have implemented the PageRank and GNMF algorithm.
The code changes required to make these algorithms run in Gilbert are minimal and only concern Gilbert's loop abstraction.
Our benchmarks have proven that Gilbert can handle data sizes which no longer can be efficiently processed on a single computer.
Moreover, Gilbert showed a promising scale out behavior, making it a suitable candidate for large-scale data processing.

The key contributions of this work include the development of a scalable data analysis tool which can be programmed using the well-known MATLAB language. We have introduced the fixpoint operator which allows to express loops in a recursive fashion. The key characteristic of this abstraction is that it allows an efficient concurrent execution unlike the \code{for} and \code{while} loops. Furthermore, we researched how to implement linear algebra operations efficiently in modern parallel data flow systems, such as Spark and Flink. Last but not least, we have shown that Flink's optimizer is able to automatically choose the best execution strategy for matrix-matrix multiplications. Even though, Gilbert can process vast amounts of data, it turned out that it cannot beat the performance of hand-tuned algorithms using Spark or Flink. This is caused by the overhead entailed by the linear algebra abstraction. The overhead is also the reason for the higher memory foot-print, causing Spark and Flink to spill faster to disk than the hand-tuned algorithms. Gilbert trades some performance off for better usability, which manifests itself in shorter and more expressive programming code. The fact that Gilbert only supports one hard-wired partitioning scheme, namely square blocks, omits possible optimization potential. Interesting aspects for further research and improvements of Gilbert include the extension of the Gilbert optimizer by new optimization strategies. The investigation of different matrix partitioning schemes and its integration into the optimizer which selects the best overall partitioning seems to be very promising. Furthermore, a tighter coupling of Gilbert's optimizer with Flink's optimizer could result in beneficial synergies. Forwarding the inferred matrix sizes to the underlying execution system might help to improve the execution plans.
